---
title: "GTEx RNAseq Deconvolution"
author: "Brian Gural"
date: "4/10/2023"
output: github_document
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
The pipeline is currently divided into 3 portions: 

**1_gtex_load.R**\
 - Loads, cleans, and saves the snRNAseq and bulk RNAseq datasets\
 - Accesses data by URL\
 - Converts AnnData file to .H5Seurat file\

**2_gtex_de_genes.R**\
 - Aggregates subject-matched single nucleus data into pseudo-bulk counts matrices, then compares them to actual bulk.\
 - Essentially, identifies differentially expressed genes between bulk and single nucleus RNAseq\
 - Can adjust several parameters including prefiltering of nuclei by RNA count/features, mitocondrial/ribosomal gene content, and doublet tags. Also can change log fold change threshold for DE and directionality (i.e. greaterAbs, lessAbs, etc.)\
 - Saves a seurat obj w/o DE genes

**3_gtex_decon.R**\
 - Preforms bulk deconvolution on several hundred GTEx samples after processing snRNAseq\
 - Clusters snRNAseq, excludes clusters on adjustable criteria (i.e. mitocondrial content, doublet scores, etc.), and repeats clustering after exclusion of nuclei\
 - Outputs stacked barcharts of composition estimates\
 
## Clustering and nuclei exclusion\
After loading our data and packages, we perform dimensional reduction and exclude low-quality clusters:

```{r load data, include = F}
setwd("/proj/raulab/users/brian/r_projects/bulk_decon")
# load libraries 
library(dplyr) # load first since we'll need dplyr::lapply()
libs <- c("Seurat", "DESeq2", "patchwork","SeuratDisk", "MuSiC", "reshape2", 
          "tidyverse", "SingleCellExperiment","harmony") # list libraries here
lapply(libs, require, character.only = T, quietly = T)

source("scripts/functions/decon_all.R") # load our own functions

##### Load data #####
gtex.sn <- LoadH5Seurat("data/processed/internal/sn_gtex_no_de.h5seurat")
gtex.bk <- read.csv("data/processed/internal/gtex_lv_counts_summed.csv", check.names = F, row.names = 2)
gtex.bk <- gtex.bk[,-1]
```

```{r cluster1.visualize, echo=TRUE, message=FALSE, warning=FALSE}

# cluster and remove mito-heavy nuclei
sn.clust <- ClusterSeurat(gtex.sn, 
                              res = 0.2,
                              subset = T,
                              min.rna.ft = 200,
                              max.rna.ft = 2500,
                              min.rna.ct = 800,
                              max.mt.pt = 0.05,
                              max.rb.pt = 0.05,
                              scrublet_score  = 0.4,
                              nfeatures = 2000,
                              harmony = T,
                              regress.by = "batch")

```

The GTEx data we're using had been pre-annotated with cell types (fine and broad), so we can overlay those labels to get an idea of what cell types we're looking at:

```{r cluster1.dim, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
# standard visuals for clusters and mitocondrial contaminants
dim.ft  <- c("Broad.cell.type", "seurat_clusters" )

plotUMAP(data = sn.clust,
         dim.ft = dim.ft,
         nrow = 2, 
         ncol = 1)
```

We can also look for poor quality clusters by overlaying mitocondrial and ribosome transcript proportions:

```{r cluster1.feat , fig.width = 7, fig.height= 4, echo=FALSE, message=FALSE, warning=FALSE}
# standard visuals for clusters and mitocondrial contaminants
feat.ft <- c("PercentMito", "PercentRibo")

plotUMAP(data = sn.clust,
         feat.ft = feat.ft,
         nrow = 1, 
         ncol = 2)
```

To quantify this, we're going to look at how many nuclei in each cluster contain more than 1% mitocondrial RNA:

```{r mito.clust, echo=FALSE, message=FALSE, warning=FALSE, fig.align= 'center'}
# select bad clusters/cells
bad.clust <- mitoProps(sn.clust,
                  cutoff = 0.01) 
mt.clust <- as.data.frame(bad.clust)
colnames(mt.clust) <- c("Seurat Cluster", "% >1% mt")
knitr::kable(mt.clust)
```

Now we'll set a cutoff off 20% of nuclei within a cluster. Once we label all of the nuclei in each mitocondrial cluster, we repeat the clustering without those nuclei.

```{r cluster2, echo=FALSE, message=FALSE, warning=FALSE, fig.align= 'center', include = F}
# select bad clusters/cells

bad.cells <- WhichCells(sn.clust, 
                        idents = names(bad.clust[bad.clust > 20])) # sets a 20% cutoff

sn.clust <- subset(gtex.sn, cells = bad.cells, invert = T)


# redo QC with removed cells 
sn.clust <- ClusterSeurat(sn.clust, 
                          res = 0.01,
                          subset = T,
                          min.rna.ft = 200,
                          max.rna.ft = 2500,
                          min.rna.ct = 800,
                          max.mt.pt = 0.05,
                          max.rb.pt = 0.05,
                          scrublet_score  = 0.4,
                          nfeatures = 2000,
                          harmony = T,
                          regress.by = "batch")

```

```{r cluster2.dim, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
# standard visuals for clusters and mitocondrial contaminants
dim.ft  <- c("Broad.cell.type", "seurat_clusters" )

plotUMAP(data = sn.clust,
         dim.ft = dim.ft,
         nrow = 2, 
         ncol = 1)
```


```{r music, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
# music_prop needs a SCE format & ExpressionSet  
#Idents(sn.clust) <- sn.clust$Broad.cell.type
sn.sce <- as.SingleCellExperiment(sn.clust, assay = "RNA")

# music_prop needs exprs matrix
bulk.es <- ExpressionSet(assayData = as.matrix(gtex.bk))
bulk.es <- exprs(bulk.es)
# estimate bulk composition with music
decon <- music_prop(bulk.mtx = bulk.es, sc.sce = sn.sce, markers = NULL,
                    clusters = "seurat_clusters", samples = "Participant.ID")

# turn music output into graph-friendly dataframe
# also adds ratios of nuclei clusters from subject-matched samples
props <- processProps(decon.obj = decon,
                           seurat.obj = sn.clust)
decon.melt = melt(props, id.vars = c("names", "group"))
colnames(decon.melt) = c('Sub',"Type", 'CellType', 'Prop')
decon.melt$CellType = factor(decon.melt$CellType, levels = unique(decon.melt$CellType))
decon.melt$Prop <- as.numeric(decon.melt$Prop)
```

```{r music.p1, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}

matches <- levels(sn.clust$Participant.ID)

# plot comparing subject-matched bulk and sn estimates

ggplot(decon.melt[which(decon.melt$Sub %in% matches),], 
       aes(x=Sub, y=Prop, fill=CellType))  +
    geom_bar(stat='identity', 
             position = "fill", 
             width = 0.5,
             color = "black")+
    scale_fill_discrete(name = "Cell Type")+
    ylab("Proportion") +
    facet_grid(~Type, scales = "free_x") +
    theme(
      axis.text.x = element_text(color = "black", size = 15),
      strip.text = element_text(size = 20),
      title = element_text(size = 20),
      legend.text = element_text(size = 18),
      axis.ticks.x = element_blank()) +
    xlab("Samples")+
    scale_x_discrete(labels = matches)

```

```{r music.p2, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}

rank <- props[order(props[which(props$group == "Bulk"),1]),length(rank)-1]


ggplot(decon.melt[which(decon.melt$Type == "Bulk"),], 
       aes(x=factor(Sub, levels = rank), y=Prop, fill=CellType))  +
  geom_bar(stat='identity', 
           position = "fill", 
           width = 1)+
  scale_fill_discrete(name = "Cell Type")+
  ylab("Proportion") +
  facet_grid(~Type, scales = "free_x") +
  theme(
    axis.text.x = element_blank(),
    strip.text = element_blank(),
    title = element_text(size = 20),
    legend.text = element_text(size = 18),
    axis.ticks.x = element_blank()) +
  xlab("GTEx Individuals")


```