---
title: "Comparison of deconvolution methods"
author: "Brian Gural"
date: "5/30/2023"
output:
  html_document:
    toc: true
    toc_float: true
---




```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction and Context\
The aim of this markdown is to take a look at how different deconvolution methods select markers and report compostition. Our single nucleus data is a merged set from two male B6 from Christoph and publically available Tabula Muris data (heart and aorta). Our bulk data is supplied by the Jensen lab, from B6 and a KO line, with heart failure (MI) and untreated (sham). Also, we're including experimentally purified and sequenced homogenous cell fractions (cardiomyocytes, endothelial cells, and fibroblasts). These fractions in duplicate from Christoph and Froese 2022 (doi: 10.1016/j.isci.2022.103965)

## Load data and packages\


```{r load libs, message=FALSE, warning=FALSE, cache=FALSE, include=F}
# load libraries
libs <- c("Seurat", "ggplot2", "DESeq2", "patchwork","SeuratDisk", "MuSiC", "reshape2",
          "tidyverse", "SingleCellExperiment","harmony", "SCpubr","shiny", 
          "AUCell", "viridis", "gplots", "scales", "ggrepel", "gridExtra", "scCustomize",
          "httr","readxl","matrixStats") # list libraries here
lapply(libs, require, character.only = T)
source("jensen/scripts/functions/decon_all.R")
```


```{r load data, include = F, cache = TRUE}
bulk <- read.csv("jensen/data/processed/jensen_rau_froese_cpm", row.names = 1)
bulk.pheno <- read.csv("jensen/data/processed/jensen_rau_froese_pheno", row.names = 1)
sn <- LoadH5Seurat("jensen/data/processed/single_cell/06062023.h5seurat")
```

## MuSiC and composition estimates\


```{r cluster2.dim, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
# music_prop needs exprs matrix
sn.sce <- as.SingleCellExperiment(sn, assay = "RNA")
bulk.es <- ExpressionSet(assayData = as.matrix(bulk))
bulk.es <- exprs(bulk.es)
# estimate bulk composition with music
# Access the cell type using the index

decon <- music_prop(bulk.mtx = 2^bulk.es, sc.sce = sn.sce, markers = NULL,
                    clusters = "ident", samples = "orig.ident")
```


```{r cluster2.feat , fig.width = 7, fig.height= 4, echo=FALSE, message=FALSE, warning=FALSE}
# turn music output into graph-friendly dataframe
# also adds ratios of nuclei clusters from subject-matched samples
decon.melt = melt(decon$Est.prop.weighted)
colnames(decon.melt) = c('Sub', 'CellType', 'Prop')
decon.melt$CellType = factor(decon.melt$CellType, levels = unique(decon.melt$CellType))

decon.melt$CellType = factor(decon.melt$CellType, levels = unique(decon.melt$CellType))
decon.melt$Prop <- as.numeric(decon.melt$Prop)

decon.melt$genotype <- "wt"
decon.melt[grepl("A", decon.melt$Sub), 4] <- "ko"
decon.melt$treatment <- "control"
decon.melt[grepl("Lx", decon.melt$Sub), 5] <- "mi"
decon.melt$type <- "whole"
decon.melt[grepl("RNA", decon.melt$Sub), 6] <- "fraction"

decon.melt$origin <- "Jensen"
decon.melt[grepl("B6", decon.melt$Sub), 4] <- "Rau"
decon.melt[grepl("Sham", decon.melt$Sub), 4] <- "Froese"

```

```{r music plot controls, fig.width = 12, fig.height= 7, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(decon.melt,
       aes(x=Sub, y=Prop, fill=CellType))  +
  geom_bar(stat='identity',
           position = "fill",
           width = 0.5,
           color = "black")+
  scale_fill_viridis_d(name = "Cell Type")+
  facet_wrap(~genotype, 
             scales = "free_x",
             labeller =  label_wrap_gen(multi_line=FALSE)) +
  ylab("Proportion") +
  theme( axis.text.x = element_text(color = "black", size = 15, angle = 30),
         strip.text = element_text(size = 20),
         title = element_text(size = 20),
         legend.text = element_text(size = 18),
         axis.ticks.x = element_blank()) +
  xlab("Samples (5%)")

```

## DirichletReg\


```{r bisque, eval=FALSE, fig.height=8, fig.width=7, message=FALSE, warning=FALSE, include=FALSE}
library("DirichletReg")
library("Ternary")

decon.mat <- decon$Est.prop.weighted
plot(decon.mat[,c(1,2,6)], cex = 0.5, a2d = list(colored = FALSE, c.grid = FALSE))


AL <- DR_data(ArcticLake[, 1:3])
# Include the full path to your data file here if necessary:
myData <- decon.mat[,c(1,2,6,7)]




plot(AL, cex = 0.5, a2d = list(colored = FALSE, c.grid = FALSE))

plot(rep(ArcticLake$depth, 3), as.numeric(AL), pch = 21, 
     bg =  rep(c("#E495A5","#86B875", "#7DB0DD"), each = 39), xlab = "Depth (m)", ylab = "Proportion", ylim = 0:1)

lake1 <- DirichReg(AL ~ depth, ArcticLake)
summary(lake1)

lake2 <- update(lake1, . ~ . + I(depth^2) | . + I(depth^2) | . + I(depth^2))
anova(lake1, lake2)
# Blood example

Bld <- BloodSamples
Bld$Smp <- DR_data(Bld[, 1:4])
blood1 <- DirichReg(Smp ~ Disease | 1, Bld, model = "alternative", base = 3)
blood2 <- DirichReg(Smp ~ Disease | Disease, Bld, model = "alternative", base = 3)

anova(blood1, blood2)

summary(blood1)

```

## Aitchinson Distance\


```{r coda, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
library("coda.base")

# make dataframe with expected compositions for each fraction (70 - 95% purity)
# rows = samples , cols = proportions
sim.samples <- bulk.pheno[which(bulk.pheno$fraction != "whole_tissue") ,c(1,3)]
cell.dict <- data.frame(bulk.pheno = unique(sim.samples$fraction),
                        cell.types = c("CMs", "fibroblasts", "endothelial cells"))
cell.types <- levels(sn@active.ident)

sim.fractions <- matrix(nrow = nrow(sim.samples), ncol = length(cell.types)) |>
  as.data.frame()
row.names(sim.fractions) <- sim.samples$id
colnames(sim.fractions) <- cell.types

purity <- 0.95
for(i in rownames(sim.fractions)){
name1 <- sim.samples[which(sim.samples$id == i),2]
major <- cell.dict[which(cell.dict$bulk.pheno == name1),2]
sim.fractions[i,major] <- purity
sim.fractions[i,cell.types[which(cell.types != major)]] <- (1 - purity) / length(cell.types)
}

# format estimates as samples by proportions
est.fractions <- decon$Est.prop.weighted[sim.samples$id,]
est.fractions[est.fractions == 0] <- 0.05 *0.65
aitch.vals <- data.frame(aitchison = rep(NA, dim(est.fractions)[1]))
                    
rownames(aitch.vals) <- rownames(est.samples)

for(i in rownames(sim.fractions)){
aitch.vals[i,] <- coda.base::dist(
                  rbind(sim.fractions[i,cell.types],est.fractions[i,cell.types]), 
                  method = 'aitchison')[1]
}
aitch.vals
```


## Bisque\
```{r DeconvoBuddies Find markers, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
# find markers
library(DeconvoBuddies)
sn <- sn.clust.2
sn.sce <- as.SingleCellExperiment(sn, assay = "RNA") |>
  as("SummarizedExperiment")
# add gene names, not sure why they broke

gene_names <- rownames(sn@assays$RNA@counts)
rownames(sn.sce) <- gene_names

# 
ratios <- get_mean_ratio2(sn.sce, cellType_col = "ident")
fc <- findMarkers_1vAll(sn.sce, cellType_col = "ident", mod = "~orig.ident")
marker_stats <- left_join(ratios, fc, by = c("gene", "cellType.target"))


cell_types <- levels(sn.sce$ident)
cell_colors <- create_cell_colors(cell_types = cell_types, pallet = "tableau", split = "\\.", preview = TRUE)


plot_marker_express(sn.sce,
    marker_stats,
    "CMs",
    n_genes = 5,
    rank_col = "rank_ratio",
    anno_col = "anno_ratio",
    color_pal = cell_colors,
    cellType_col = "ident"
)


# Select top 100 markers for each cellType.target
top_markers <- marker_stats %>%
  group_by(cellType.target) %>%
  top_n(-500, rank_marker) %>%
  ungroup()
```

```{r bisque, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}


markers <- top_markers[,c(1,2,9)] |>
  as.data.frame()

colnames(markers)[c(2,3)] <- c("cluster","avg_logFC")

bulk.eset <- Biobase::ExpressionSet(assayData = as.matrix(2^bulk))

bisque.marker <- MarkerBasedDecomposition(bulk.eset, markers, weighted = T)
library(Biobase)
library(BisqueRNA)
# music_prop needs exprs matrix
sample.ids <- colnames(sn)

# make expression sets
sc.pheno <- data.frame(check.names=F, check.rows=F,
                       stringsAsFactors=F,
                       row.names=sample.ids,
                       SubjectName=sn$orig.ident,
                       cellType=sn@active.ident)
sc.meta <- data.frame(labelDescription=c("SubjectName",
                                         "cellType"),
                      row.names=c("SubjectName",
                                  "cellType"))
sc.pdata <- new("AnnotatedDataFrame",
                data=sc.pheno,
                varMetadata=sc.meta)
sc.eset <- Biobase::ExpressionSet(assayData= as.matrix(sn@assays$RNA@counts),
                                  phenoData=sc.pdata)
# Bisque decon
res <- BisqueRNA::ReferenceBasedDecomposition(bulk.eset, sc.eset, markers=NULL, use.overlap=FALSE)

```

```{r bisque process , fig.width = 7, fig.height= 4, echo=FALSE, message=FALSE, warning=FALSE}
# turn music output into graph-friendly dataframe
# also adds ratios of nuclei clusters from subject-matched samples
decon.melt <- melt(res$bulk.props)
decon.melt <- decon.melt[,c(2,1,3)]
colnames(decon.melt) = c('Sub', 'CellType', 'Prop')
decon.melt$CellType = factor(decon.melt$CellType, levels = unique(decon.melt$CellType))

decon.melt$CellType = factor(decon.melt$CellType, levels = unique(decon.melt$CellType))
decon.melt$Prop <- as.numeric(decon.melt$Prop)

decon.melt$genotype <- "wt"
decon.melt[grepl("A", decon.melt$Sub), 4] <- "ko"
decon.melt$treatment <- "control"
decon.melt[grepl("Lx", decon.melt$Sub), 5] <- "mi"
decon.melt$type <- "whole"
decon.melt[grepl("RNA", decon.melt$Sub), 6] <- "fraction"

decon.melt$origin <- "Jensen"
decon.melt[grepl("B6", decon.melt$Sub), 4] <- "Rau"
decon.melt[grepl("Sham", decon.melt$Sub), 4] <- "Froese"

```

```{r bisque plot controls, fig.width = 12, fig.height= 7, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(decon.melt,
       aes(x=Sub, y=Prop, fill=CellType))  +
  geom_bar(stat='identity',
           position = "fill",
           width = 0.5,
           color = "black")+
  scale_fill_viridis_d(name = "Cell Type")+
  facet_wrap(~genotype, 
             scales = "free_x",
             labeller =  label_wrap_gen(multi_line=FALSE)) +
  ylab("Proportion") +
  theme( axis.text.x = element_text(color = "black", size = 10, angle = 30),
         strip.text = element_text(size = 20),
         title = element_text(size = 20),
         legend.text = element_text(size = 18),
         axis.ticks.x = element_blank())

```

```{r dwls , fig.width = 7, fig.height= 4, echo=FALSE, message=FALSE, warning=FALSE}
library(DWLS)

dwls.sig <- buildSignatureMatrixUsingSeurat(sn.clust.2@assays$RNA@counts, Idents(sn.clust.2), "jensen/results/dwls")
```