---
title: "Merge scRNAseq datasets for deconvolution"
author: "Brian Gural"
date: "2023-08-16"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, message = F, warning = F, error = T)
```

```{r load libs, message=FALSE, echo = T, warning=FALSE, results= 'hide', cache=FALSE}
libs <- c("Seurat", "ggplot2", "DESeq2", "patchwork","SeuratDisk", "MuSiC", "reshape2",
          "tidyverse", "SingleCellExperiment", "harmony", "SCpubr", 
          "AUCell", "viridis", "gplots", "scales", "ggrepel", "gridExtra", "scCustomize",
          "httr","matrixStats", "scran", "scuttle", "scater", "DropletUtils", "scDblFinder",
          "DeconvoBuddies", "readr", "SingleCellExperiment",
          "SummarizedExperiment", "Biobase", "ggmagnify", "stringr", "MuSiC",
          "ggforce", "DirichletReg", "readxl") # list libraries here
lapply(libs, require, character.only = T)
source("jensen/scripts/functions/decon_all.R")
rm(libs)
```


```{r load data, message=FALSE, echo = T, warning = FALSE, cache = T, collapse = T}
# load individual seurats
dir <- "jensen/data/processed/single_cell/no_doublets/08162023/"
files <- list.files(dir)

sn <- lapply(files, function(x){LoadH5Seurat(paste0(dir, x))})

names(sn) <- lapply(strsplit(files, ".h5seurat"), "[[", 1)

# Load bulk data
bulk <- read_csv("jensen/data/processed/jensen_rau_froese_cpm") |>
  column_to_rownames(var = "...1")

# Load and format bulk data
jensen_bulk <- readxl::read_xlsx("jensen/data/raw/jensen_counts_correct.xlsx")[,-c(20,21)] |>
  as.data.frame()
colnames(jensen_bulk)[c(20,21)] <- c("WT MI", "KO MI")

```

```{r rau processing, fig.width=5}
# subset to the rau data
sn.rau <- sn[str_detect(names(sn), "rau")]
sn.rau <- merge(sn.rau[[1]], sn.rau[[2]])

sn <- sn[!str_detect(names(sn), "rau")]
names <- names(sn)
sn <- c(sn, sn.rau)
names(sn) <- c(names, "rau")

rm(sn.rau)
rm(names)
gc()
```


```{r non rau processing}
sn$wu_2021$PercentMito <- PercentageFeatureSet(sn$wu_2021, pattern = "^mt-")
sn$wu_2021$PercentRibo <- PercentageFeatureSet(sn$wu_2021, pattern = "^Rpl|^Rps")
```


```{r raw processing}
sn <- lapply(sn, function(x){ClusterSeurat(x, max.rna.ft = 6000, subset = T, max.mt.pt = 5, harmony = F)})

# Add cluster info

.addClusters <- function(sn.func){
  sn.func$seurat.clusters <- Idents(sn.func)
  return(sn.func)
}

sn <- lapply(sn, function(x){.addClusters(x)})
```


```{r sn raw visuals, fig.height=22, fig.width=14}
hist(sn$martini_2019$nCount_RNA, breaks = 100)
hist(sn$martini_2019$nFeature_RNA, breaks = 100)
# umap
# PercentMito
# nCount RNA
# nFeature RNA
# doublet score
.getPlots <- function(sn.func, names){
  dim.ft  <- c("seurat.clusters")
  feat.ft <- c("nCount_RNA","PercentMito", "DoubletScore")
  vln.ft <- c("PercentMito", "DoubletScore")
  vln.groups <- c("seurat.clusters")
  plots.func <- plotUMAP(
         data = sn.func,
         dim.ft = dim.ft,
         feat.ft = feat.ft,
         vln.ft = NULL,
         vln.groups = NULL,
         ncol =1,
         font.size = 14,
         name = names)
  return(plots.func)
}

raw.plots <- lapply(names(sn), function(x){.getPlots(sn[[x]], names = x)})

names(raw.plots) <- names(sn)
.wrapPlots <- function(sn.name, plots.func){
    wrap_plots(plots.func[[sn.name]], ncol = 1 ) + 
    plot_annotation(title = sn.name,
                    theme = theme(plot.title = element_text(size = 18,
                                                            hjust = 0.5)))
}

wrapped.plots <- lapply(names(raw.plots), function(x){.wrapPlots(x, raw.plots)})

wrap_plots(wrapped.plots, ncol = length(raw.plots)) # +
  #plot_annotation(tag_levels = list(c("Clusters", "Total Counts", "Percent\nMt RNA", "Doublet Score")), theme = theme(plot.tag = element_text(size = 26), plot.tag.position = "right"))


```


```{r merge}

sn.merge <- merge(sn[[1]], sn[[2]]) |>
            merge(sn[[3]]) |>
            merge(sn[[4]])  |>
            subset(isDoublet == "no")

sn.merge <- ClusterSeurat(sn.merge, subset = F, regress.by = "origin", res = 0.01)
```

```{r merge visual}
do_DimPlot(sn.merge, group.by ="origin")
```

```{r AUCell markers, echo=FALSE,warning=FALSE, include=TRUE, fig.height = 8, fig.width = 12}
# Annotate to cell types
markers.broad <- read.csv("jensen/data/processed/external/mclellan_2020/mclellan_cell_markers_broad.csv")
```

As we talked about last time, AUCell assigns an AUC score (think enrichment) for each cell based on expression and cell types with clear bimodal distributions are retained. The second part, retaining only cell-types with distinct groups, is a change from the last workflow that has been very helpful. Here’s how those AUC distributions look for the cell types we retained:

```{r AUCell plot, echo=FALSE,warning=FALSE, include=TRUE, fig.height = 18, fig.width = 12}
# Annotate to cell types
n_markers <- 25
n_cells <- 15000
 
  seurat <- sn.merge
  cells <- sample(colnames(sn.merge), n_cells, replace = FALSE)
  seurat <- seurat |>
    subset(cells = cells)
  markers.sub <- markers.broad %>%
    group_by(cluster) %>%
    top_n(n_markers, gene) |>
    as.data.frame()
  # Split 'markers' into separate data frames for each unique subcluster
  subclusters <- split(markers.sub, markers.sub$cluster)
  # Extract the gene names from each subcluster data frame
  subcluster_genes <- lapply(subclusters, function(x) x$gene)
  # Create a list with named elements corresponding to each subcluster
  geneSets <- setNames(subcluster_genes, names(subclusters))
  # make expression matrix of single cell
  my.expr <-  seurat@assays$RNA@counts
  # Find  AUC for each cell by each type
  cells_AUC <- AUCell_run(my.expr, geneSets)
  
  cells_assignment <- AUCell_exploreThresholds(cells_AUC, plotHist=F, assign=TRUE)
  
  comments <- lapply(names(cells_assignment), function(cell_type){cells_assignment[[cell_type]]$aucThr$comment})
  good_cells <- names(cells_assignment) #[comments == ""]
  # get the thresholds and report when/how they're passdd
  selectedThresholds <- getThresholdSelected(cells_assignment)
  
  auc.val <- cells_AUC@assays@data@listData$AUC |>
    t()|>
    as.data.frame() 
  auc.val <- auc.val[,good_cells]
  auc.val$seurat <- seurat$seurat_clusters[rownames(auc.val)]

```

```{r plot heatmap, fig.height = 20}
# plot 
plotAUC <- function(cell_type, auc.df, thresholds){
threshold <- thresholds[[cell_type]]
# Create the histogam
ggplot(auc.df, aes(x = !!sym(cell_type))) +
  geom_histogram(aes(fill = !!sym(cell_type) > threshold), binwidth = 0.01, stat = "bin") +
  scale_fill_manual(values = c("TRUE" = "#D95F02", "FALSE" = "#1B9E77"), guide = FALSE) +
  geom_vline(aes(xintercept = threshold), linetype="dashed", color = "black") +
  theme_minimal() +
  theme(title = element_text(size = 18),
        axis.text = element_text(size = 14)) +
  labs(x = "Value", y = "Count", title = cell_type, subtitle = paste0("Threshold: ", round(threshold, 3))) 
}


plots.auc <- lapply(good_cells, function(x){plotAUC(cell_type = x, auc.df = auc.val, thresholds = selectedThresholds)})
 
wrap_plots(plots.auc, ncol = 2)
```

```{r AUCell heatmap, echo=FALSE,warning=FALSE, include=TRUE, fig.height = 8, fig.width = 12}
# Make correlation matrix for Seurat clusters and cell correlations to cell types

corrs <- model.matrix(~0 + ., data = auc.val) |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame()
# keep seurat cluster names just as a number
corrs <- corrs[colnames(corrs)[!grepl("seurat", colnames(corrs))], colnames(corrs)[grepl("seurat", colnames(corrs))]] 
colnames(corrs) <- lapply(strsplit(colnames(corrs), "seurat"), "[[", 2)

# format for plotting
corrs <- data.matrix(corrs, rownames.force = T)
empty_columns <- colSums(is.na(corrs) | corrs == "") == nrow(corrs)
standardized_residuals <- corrs[, !empty_columns]

heatmap.2(standardized_residuals, 
          trace = "none", 
          col = viridis(8),
          main = "Correlation coeffiecients between cell-types and clusters",
          xlab = "Seurat Clusters",
          ylab = "Cell Types",
          margins = c(8, 12),
          key.title = "Correlations",
          srtCol = 0,
          cexRow = 1.2)

Idents(sn.merge) <- sn.merge$seurat_clusters

sn.merge <- AssignAndFilterClusters(sn.merge, res.thresh = 0.25, ratio.thresh = 1, min.cell = 200)

```


```{r plot cell types, fig.height = 20, fig.width = 18}

plots <- c()
# seurat after 1_1_jensen_doublets.R and 1_2_jensen_merge.R
plots[["dim.clust"]] <- do_DimPlot(sn.merge, label = T, repel = T) + NoLegend()
plots[["dim.origin"]] <- do_DimPlot(sn.merge, group.by = "origin", label = T, repel = T) + NoLegend()

# Convert your table to a data frame and reset row names
df <- as.data.frame(table(sn.merge$origin, Idents(sn.merge)))
colnames(df) <- c("origin", "cluster", "count")

# Convert data frame from wide format to long format

# Convert cluster to factor for ordered x-axis
df$cluster <- factor(df$cluster, levels = unique(df$cluster))

# Create the bar plot
plots[["bar.clust.orig"]] <- ggplot(df, aes(x = cluster, y = count, fill = origin)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Cluster", y = "Count", fill = "Origin", 
       title = "Count of Origins by Cluster") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plots[["bar.umi"]] <- VlnPlot(sn.merge, group.by = "origin", features = "nCount_RNA", log = T) 

plots[["feat.count"]] <- do_FeaturePlot(sn.merge, features = "nCount_RNA") 
plots[["feat.features"]] <- do_FeaturePlot(sn.merge, features = "nFeature_RNA") 
plots[["feat.mito"]] <- do_FeaturePlot(sn.merge, features = "PercentMito") 


wrap_plots(plots, ncol = 2)

```
```{r DeconvoBuddies Find markers, fig.width = 7, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
# Update sn.merge with annotations
sn.merge <- sn.merge |>
         subset(features = rownames(sn.merge)[rownames(sn.merge) %in% rownames(bulk)]) 

sn.sce <- sn.merge|> 
  as.SingleCellExperiment(assay = "RNA") |>
  as("SummarizedExperiment")

# Add gene names
rownames(sn.sce) <- rownames(sn.merge@assays$RNA@counts)

# Get markers based on expression ratios between clusters and 1vAll comparisons

scn.markers  <- scran::findMarkers(sn.sce, groups = Idents(sn.merge), pval.type = "all", direction = "up")


.getMarkers <- function(type){
marker <- scn.markers@listData[[type]] |>
  as.data.frame() |> 
  subset(p.value == 0) |>
  arrange(desc(summary.logFC)) |>
  slice_head(n = 25) |>
  select(p.value, FDR, summary.logFC) |>
  mutate(celltype = type) |> 
  rownames_to_column(var = "gene")
return(marker)}


all.markers <- lapply(levels(Idents(sn.merge)), function(x){.getMarkers(x)}) |> 
               reduce(full_join)


gc()
```

```{r deconvolute music , echo=FALSE, warning=FALSE, include=T}
# Prep bulk matrix as ExpressionSet
bulk.es.exp <- bulk |> 
  as.matrix() %>%
  ExpressionSet(assayData = .) |> 
  exprs()

# Run MuSiC with subset genes
decon.melt <- EstimateCellTypeProportions(sn.merge, bulk.es.exp, for.aitchison = FALSE, bulk.log2 = TRUE,
                                          sn.individuals = "orig.ident", marker = all.markers$gene, cells_exclude = NULL)
 
```

```{r prep phenos , fig.width = 14, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
phenotypes_real <- jensen_bulk[1,] |> 
  pivot_longer(cols = -c(1:5), 
               names_to = "gene_treat", 
               values_to = "id") |>
  mutate(gene_treat = str_remove(gene_treat, "\\..*")) |>
  select(gene_treat, id)

```

                              
```{r prep phenos , fig.width = 14, fig.height= 8, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)

# Turn music output into graph-friendly dataframe
decon.melt <- decon.melt |>
  mutate(
    CellType = factor(CellType, levels = unique(CellType)),
    Sub = factor(Sub, levels = unique(Sub)),
    Prop = as.numeric(Prop)
  )

# Bring in real phenotype data
decon.melt <- merge(decon.melt, phenotypes_real, by.x = "Sub", by.y = "id", all.x = TRUE) |>
  mutate(
    gene_treat = factor(gene_treat, levels = unique(gene_treat))
  )

# Add genotype info
decon.melt <- decon.melt |>
  mutate(
    genotype = factor(ifelse(!is.na(gene_treat), str_split(gene_treat, " ", simplify = TRUE)[,1], NA)),
    treatment = factor(ifelse(!is.na(gene_treat), str_split(gene_treat, " ", simplify = TRUE)[,2], NA)),
    type = factor(ifelse(str_detect(Sub, "RNA|Sham"), "fraction", "whole")),
    origin = factor(case_when(
      str_detect(Sub, "B6") ~ "Rau",
      str_detect(Sub, "Sham") ~ "Froese",
      TRUE ~ "Jensen"
    )),
    fraction_type = factor(case_when(
      str_detect(Sub, "CM") ~ "Cardiomyocytes",
      str_detect(Sub, "Fib|FB") ~ "Fibroblasts",
      str_detect(Sub, "Endo|EC_") ~ "Endothelial Cells",
      TRUE ~ NA_character_
    ))
  )





```

```{r music plot sham, fig.width = 19, fig.height= 9, echo=FALSE, message=FALSE, warning=FALSE}

# Reorder the factor levels of CellType so that they are ordered by sum of their values accross groups
decon.melt$CellType <- reorder(decon.melt$CellType, -decon.melt$Prop, FUN = sum)

# Setting the levels
decon.melt$gene_treat <- factor(decon.melt$gene_treat, 
                                        levels = c('WT sham', 'WT MI', 'KO sham', 'KO MI'))

#! These colors are hard to read when printed, should be lighter
#! Or the dots from geom_jitter need to be bigger/brighter

my_palette <- c('#78c679', '#006d2c', '#d95f0e', '#993404')

plot.jen.sample <- decon.melt |>
  subset(type == "whole") |>
  ggplot(aes(x = as.numeric(CellType), y = Prop, fill = gene_treat)) +
  geom_bar(stat = "summary", 
           position = position_dodge(0.9), 
           fun = mean,
           width = 0.9,
           color = "black") +
  theme(axis.text.x = element_text(color = "black", size = 25, angle =15),
        axis.text.y = element_text(color = "black", size = 20),
        axis.ticks = element_blank(),
        legend.position = c(0.95, 0.95),
        legend.justification = c("right", "top"),
        legend.box.just = "right",
        legend.margin = margin(6, 6, 6, 6),
        strip.text = element_text(size = 20),
        title = element_text(size = 30),
        legend.text = element_text(size = 25),
         panel.background = element_rect(fill='transparent'),
         plot.background = element_rect(fill='transparent', color=NA),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         legend.background = element_rect(fill='transparent'),
         legend.box.background = element_rect(fill='transparent')) +
  labs(x = "Cell Type", 
       y = "Proportion", 
       fill = "Genotype-Treatment") +
  scale_fill_manual(values = my_palette) +  # using custom palette
  geom_jitter(inherit.aes = T, 
              position = position_jitterdodge(0.01, 0, 0.9),
              size = 2) +
  scale_x_continuous(
    breaks = 1:length(levels(decon.melt$CellType)),
    label = levels(decon.melt$CellType)) +
  geom_magnify(from = c(2.55,3.45,0,0.2), to = c(1.7,4,0.33,0.8), 
                   shape = "rect", shadow = TRUE) 

plot.jen.sample
```

```{r bcvs.fractions, fig.width = 16, fig.height= 4, echo=FALSE, message=FALSE, warning=FALSE}

# Plot the samples which are purified, bulk sequenced cell types
plot.fractions <- decon.melt |>
  subset(type == "fraction") |>
  ggplot(aes(x=Sub, y=Prop, fill=CellType))  +
   geom_bar(stat='identity',
           position = "fill",
           width = 0.8,
           color = "black")+
   scale_color_brewer(palette = "dark2")  +
   facet_wrap(~fraction_type, 
           scales = "free_x",
           labeller =  label_wrap_gen(multi_line=FALSE)) +
   ylab("Proportion") +
   theme_minimal() +
   theme(axis.text.x = element_blank(),
        axis.text.y = element_text(color = "black", size = 20),
        axis.ticks = element_blank(),
        strip.text = element_text(size = 30),
        title = element_text(size = 30),
        legend.text = element_text(size = 25),
        panel.background = element_blank(),
        panel.grid = element_blank()) +
   xlab("Sample replicates") 

plot.fractions

```

```{r dirichlet, fig.width = 12, fig.height= 6, echo=TRUE, message=TRUE, warning=FALSE}

## Prep DirichletReg matrix

# Add small value to remove zeros
decon.melt$Prop <- decon.melt$Prop + 0.0001

# Make model matrix
dir.mat <- decon.melt |>
  subset(type == "whole",
         select = c("Sub", "CellType", "Prop", "genotype", "treatment")) |>
  dcast(Sub + genotype + treatment ~ CellType, value.var = "Prop")


### TEMP ###

############

# Convert data to DirichletRegData object
dir.mat$CellTypes <- DR_data(dir.mat[,c(4:length(dir.mat))])

dir.mat$genotype   <- as.factor(dir.mat$genotype) |>
                        relevel(ref = "WT") 

dir.mat$treatment <- as.factor(dir.mat$treatment) |>
                        relevel(ref = "sham") 
# Run Dirichlet regression
model.1 <- DirichReg(CellTypes ~ treatment * genotype, data = dir.mat, model = "alternative", base = 3)
model.2 <- DirichReg(CellTypes ~ treatment * genotype, data = dir.mat, model = "common")

# compare models, find if interaction term improves model

summary(model.2)
model.1$sub.comp

```

